---
title: "Hanover_LandUse"
output: html_document
date: '2023-05-23'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("sf")
#install.packages("leaflet")
#install.packages("writexl")
#install.packages("colorspace") 
#install.packages("RColorBrewer")
library(sf)
library(leaflet)
library(dplyr)
library(readxl)
library("writexl")
library(colorspace) 
library(RColorBrewer)
library(ggplot2)
library(plotly)
library(viridis)
library(viridisLite)
```

* Set working directory. SAVE DATA FILES FROM TEAMS, CHANGE WD TO LOCATION YOU SAVE THEM AND ALL OTHER CODE SHOULD WORK 
```{r wd}
#setwd("C:/Users/ariadnetynes/Desktop/VT_DSPG/DSPG2023_VCE_Hanover")
```

# read in shapefile (most up to date version with some variables)
```{r wd}
shapefile <- st_read("Hanover_Parcels.shp")
```

# read in excel sheet (more parcel variables - tax assessed data)
```{r wd}
assess <- read_excel("Assessor_Data.xlsx")
```

# Create a leaflet map of shapefile 
```{r wd}
leaflet() %>%
  addTiles() %>%
  addPolygons(data = shapefile, weight = 0.1)  # Add the shapefile to the map
```

# summarize zoning variable from shapefile 
```{r sum}
# Convert the vector to a factor (categorical variable)
categorical_var <- shapefile$ZONING_LIS
# Summarize the categorical variable
summary_stats <- table(categorical_var)
# View the summary statistics
print(summary_stats)
```

# summarize zoning variable from assessor data  
```{r sum}
# Convert the vector to a factor (categorical variable)
categorical_var <- assess$Zone
# Summarize the categorical variable
summary_stats <- table(categorical_var)
# View the summary statistics
print(summary_stats)
```


# Merge the two data sets to get one huge data set 
```{r }
merged_data <- merge(shapefile, assess, by.x = "GPIN", by.y = "GPIN")

# df data set shows the 4,631 parcels who do not have the same zone
fik <- merged_data[merged_data$Zone != merged_data$ZONING_LIS, ]
df <- fik[, c("Zone", "ZONING_LIS")]
```

#create combined variable for zone
```{r }
#dataset$newvariable <- ifelse(condition, if true, if false)
#condition = zoning variable is the same in both data sets
# if true = use the variable, if false = "NULL"
merged_data$comb_zone <- ifelse(merged_data$Zone == merged_data$ZONING_LIS, merged_data$Zone, "NULL")

#dataset$newvariable <- ifelse(condition, if true, if false)
#condition = zoning variable is "See Map" from most up-to-date, "shapefile" data set 
# if true = use the zoning variable from the assessor excel file
# if false = use most recent data set "shapefile"
merged_data$comb_zone <- ifelse(merged_data$ZONING_LIS == "See Map", merged_data$Zone, merged_data$ZONING_LIS)

# Specify the variables to relocate
variables_to_relocate <- c("GPIN", "comb_zone", "LOT_ACRES", "LAND_VALUE", "LEGALDESC")
# Relocate the variables - move important variables to front of data set 
merged_data <- merged_data %>%
  select(one_of(variables_to_relocate), everything())
```



# Summary stats for new, cleaned variable (comb_zone)
```{r sum}
allzones <- merged_data[, c("comb_zone", "Zone", "ZONING_LIS")]

# Convert the vector to a factor (categorical variable)
categorical_var <- merged_data$comb_zone
# Summarize the categorical variable
summary_stats <- table(categorical_var)
# View the summary statistics
print(summary_stats)
```


# create a variable for landuse - residential, commercial, industrial, agricultural, conservation 

```{r }
land_use <- function(comb_zone) {
  if (is.na(comb_zone)) {
    return("Missing")
  } else if (comb_zone == "RC" || comb_zone == "RR-C") {
    return("Conservation")
  } else if (comb_zone == "AR-1" || comb_zone == "AR-2" || comb_zone == "AR-3") {
    return("Agriculture/Residential")
  } else if (comb_zone == "A-1") {
    return("Agriculture")
  } else if (comb_zone == "R-1" || comb_zone == "R-2" || comb_zone == "R-3" || comb_zone == "R-4" || comb_zone == "R-5" || comb_zone == "R-6" || comb_zone == "RS" || comb_zone == "RM") {
    return("Residential")
  } else if (comb_zone == "B-1" || comb_zone == "B-2" || comb_zone == "B-20" || comb_zone == "B-4" || comb_zone == "B-3" || comb_zone == "B-0" || comb_zone == "RS" || comb_zone == "RM") {
    return("Commercial")
  } else if (comb_zone == "M-1" || comb_zone == "M-2" || comb_zone == "M-3") {
    return("Industrial")
  } else {
    return("Other")
  }
}

merged_data$land_use <- sapply(merged_data$comb_zone, land_use)

```

```{r sum}
allzones <- merged_data[, c("land_use", "comb_zone", "Zone", "ZONING_LIS")]

# Convert the vector to a factor (categorical variable)
categorical_var <- merged_data$land_use
# Summarize the categorical variable
summary_stats <- table(categorical_var)
# View the summary statistics
print(summary_stats)
```


# Create a leaflet with parcels colored by land use variable
```{r}

color_map <- colorFactor(palette = viridis(length(unique(merged_data$land_use))), 
                         domain = merged_data$land_use)

leaflet() %>%
  addTiles() %>%
  addPolygons(data = merged_data, weight = 0.1, smoothFactor = 0.2, fillColor = ~color_map(land_use))

```

```{r}
num_cat <- 34
#custom_pal <- brewer.pal(num_cat, "viridis")
#custom_pal <- brewer.pal(8, "Set2")
custom_pal <- viridis(34)
categorical_var <- as.factor((categorical_var))
custom_zoneCol <- setNames(custom_pal, levels(categorical_var))
zoneCol <- colorFactor(custom_zoneCol, levels = levels(categorical_var))

zone_plot <- leaflet(data = merged_data) %>%
  addTiles() %>%
  addPolygons(color = ~zoneCol(categorical_var), weight = 0.2, stroke = FALSE, smoothFactor = 0.2, fillOpacity = 1,label = paste("Land Type: " , merged_data$land_use), highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE)) %>%
  addLegend(position = "bottomright", pal = zoneCol, values = categorical_var, opacity = .8, title = "Categories of Zones") 

zone_plot
```

#Create descriptive stats and graphs
```{r Lot Acres}
acres_landuse <- merged_data %>%
  select(LOT_ACRES, land_use)
  
acres_landuse$land_use <- as.factor(acres_landuse$land_use)
ggplot(acres_landuse, aes(land_use, LOT_ACRES, color = land_use))+
  geom_col()+
  theme(legend.position = "none", axis.text.x = element_text(size = 7))+
  labs(x = "Land Use", y="Lot Acres")+
  scale_y_continuous(labels = scales::comma)+
  ggtitle("Lot Acres By Land Use")

acre <- as.data.frame(acres_landuse) %>%
  na.omit()%>%
  group_by(land_use)%>%
  summarize(mean_acres = mean(LOT_ACRES),
    median_acres = median(LOT_ACRES),
    min_acres = min(LOT_ACRES),
    max_acres = max(LOT_ACRES),
    sd_acres = sd(LOT_ACRES))

acre

mean_a <-ggplot(acre, aes(land_use, mean_acres, fill = land_use))+
  geom_col()+
  scale_fill_viridis_d()+
  theme(legend.position = "none", axis.text.x = element_text(size = 7))+
  labs(x = "Land Use Type", y="Lot Acres", caption = "Data Source: Hanover County Assessor Data")+
  ggtitle("Mean Lot Acres By Land Use Type")

#mean_a <-ggplotly(mean_a, tooltip = "y")
mean_a

```

```{r Year Built}
yr <- merged_data[merged_data$YEAR_BUILT != 0, ]
df_year <- yr[, c("YEAR_BUILT", "land_use")]

ggplot(df_year, aes(YEAR_BUILT, land_use, color = land_use))+
  geom_point()

ggplot(df_year, aes(YEAR_BUILT, fill = land_use))+
  geom_histogram()

ggplot(df_year, aes(YEAR_BUILT, land_use, fill = land_use))+
  geom_col()

#year_landuse <- merged_data %>%
#  select(df_year, land_use)
  


#year_landuse <- merged_data %>%
 # select(YEAR_BUILT, land_use){
  #  if 
  #}
  

#mybins <- c(0, seq(1800, 2100, by = 1000)) 
  
#year_landuse$land_use <- as.factor(year_landuse$land_use)
#ggplot(year_landuse, aes(YEAR_BUILT, land_use))+
 # geom_point()


#as.data.frame(year_landuse) %>%
#  na.omit()%>%
#  group_by(YEAR_BUILT)%>%
#  summarize(mean_acres = mean(land_use),
#    median_acres = median(land_use),
#    min_acres = min(land_use),
#    max_acres = max(land_use),
#    sd_acres = sd(land_use))

```

```{r Land Value}
values <- merged_data %>%
  select(LANDVALUE, TOTALVALUE, land_use)
  
#values$TOTALVALUE <- as.factor(values$TOTALVALUE)
#ggplot(values, aes(x = land_use,y = LANDVALUE, fill = as.factor(TOTALVALUE)))+
#  geom_col()+
#  position_dodge(width= 0.5)
#  labs(x = "Land Use", y = "Land Value", fill = "Total Value")

 

vall <- as.data.frame(values) %>%
  na.omit()%>%
  group_by(land_use)%>%
  summarize(mean_landval = mean(LANDVALUE),
    median_landval = median(LANDVALUE),
    min_landval = min(LANDVALUE),
    max_landval = max(LANDVALUE),
    sd_landval = sd(LANDVALUE))

vall2 <- as.data.frame(values) %>%
  na.omit()%>%
  group_by(land_use)%>%
  summarize(mean_totval = mean(TOTALVALUE),
    median_totval = median(TOTALVALUE),
    min_totval = min(TOTALVALUE),
    max_totval = max(TOTALVALUE),
    sd_totval = sd(TOTALVALUE))

#ggplot(vall, aes(land_use, mean_landval, fill = land_use))+
#  geom_col()+
#  theme(legend.position = "none", axis.text.x = element_text(size = 7))+
#  labs(x = "Land Use", y="Land Value")+
#  scale_y_continuous(breaks = seq(0, 200000, by = 50000))
#  ggtitle("Mean Land Value")
vals_lmean<- ggplot(vall, aes(land_use, mean_landval, fill = land_use))+
  geom_col()+
  scale_fill_viridis_d()+
  theme(legend.position = "none", axis.text.x = element_text(size = 7))+
  labs(x = "Land Use", y="Land Value", caption = "Data Source: Hanover County Assessor Data")+
  scale_y_continuous(labels = scales::comma)+
  ggtitle("Average Land Value by Land Type")

vals_tmean <- ggplot(vall2, aes(land_use, mean_totval, fill = land_use))+
  geom_col()+
  scale_fill_viridis_d()+
  theme(legend.position = "none", axis.text.x = element_text(size = 7))+
  labs(x = "Land Use Type", y="Total Value", caption = "Data Source: Hanover County Assessor Data")+
  scale_y_continuous(labels = scales::comma)+
  ggtitle("Average Parcel Value by Land Type")

#vals_lmean <-ggplotly(vals_lmean, tooltip = "y")
#vals_tmean <-ggplotly(vals_tmean, tooltip = "y")

vals_lmean
vals_tmean
```




